Freezing of a Weeks-Chandler-Anderson Fluid using Classical Density Functional Theory

### 2021-01-15 papers/thesis-kirstie/thesis.tex
edits from david

A description of cDFT often begins with considering the Grand Free 
Energy defined as
\begin{equation}\Phi=F-\mu{N}\end{equation}
which can then be expressed in terms of a varying number density 
profile $n(\vec r)$
\begin{equation}\label{GrandFE}\Phi= F_{int} +\int n(\vec{r})\phi_{ext}{(\vec r)}d\vec{r}-\mu\int n(\vec r)d\vec{r}\end{equation}
where $\phi_{ext}=\frac{V_{ext}}{N}$ is an externally applied potential energy 
$V_{ext}(\vec r)$ per particle, an intrinsic Helmholtz free energy 
is defined as
\begin{equation}F_{int} = F - \int n(\vec{r})\phi_{ext}{(\vec r)}d\vec{r}\end{equation}  
and 
\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation}%
The Grand free energy is applicable to systems with a varying number of 
particles and fixed chemical potential, temperature and volume. But in 
this paper a system with a fixed temperature, volume, and number of 
particles will be of interest in which case
%\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation} and so
the Grand free energy is constrained
\begin{equation}\label{GrandFE}\Phi_{constrained}= F[n(\vec r)]-\mu N\end{equation}
The dependence on $n(\vec r)$ is now only present in the expression for 
the Helmholtz free energy, and so it is the Helmholtz energy as a functional 
of the number density profile $F[n(\vec r)]$ that is minimized at equilibrium. 
More specifically, in this paper, the Helmholtz free energy per particle, 
or atom,  $f_N=\frac{F}{N}$ will be minimized which is given by
\begin{equation}f_N=\frac{U}{N}-\frac{TS}{N}\end{equation}
with differential
\begin{equation}\label{usetoshowmin}df_N=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
%This result shows that the natural variables of the Helmholtz free energy 
%per particle $f_N$ are temperature, volume and one over the number of 
%particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. 
%\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}
%Taking the differential
which can be set equal to 
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}\bigg|_{V,\frac{1}{N}}dT 
+\frac{\partial f_N}{V}\bigg|_{T,\frac{1}{N}}dV + \frac{\partial f_N}
{\partial \left(\frac{1}{N}\right)}\bigg|_{T,V}d\left(\frac{1}{N}\right)\end{displaymath}
Using the relation
\begin{displaymath}d\left(\frac{1}{n}\right)=Vd\left(\frac{1}{N}\right)+\frac{1}{N}dV\end{displaymath}
we get
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}\bigg|_{\frac{1}{n}}dT 
+ \frac{\partial f_N}{\partial \frac{1}{n}}\bigg|_{T}d\left(\frac{1}{n}\right)\end{displaymath}
\begin{equation}f_N(T,\frac{1}{n})\end{equation}
This result shows that temperature and the inverse number density, 
or volume per particle, are natural variables of the Helmholtz free energy 
per particle $f_N$.
%which will thus be minimized when these natural variables are held fixed.

@@

The Helmholtz free energy can be written as
\begin{equation}{F=-k_{B}T\ln(Z)}\end{equation}
which results from equating the expressions for the Helmholtz free energy 
from thermodynamics to those of statistical mechanics. The paritian function 
Z is integrated over all of phase space for N identical particles each with 
a momentum $\vec{p}$ and a position $\vec{r}$, and is given by
\begin{align}  \label{Z_total}
     Z=\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}\int{dQ}~e^\frac{-H(\vec{p}_1,\vec{p}_2,...\vec{p}_N,\vec{r}_1, \vec{r}_2,...\vec{r}_N)}{k_BT}
\end{align}
where
\begin{displaymath}{dP=d\vec{p}_1d\vec{p}_2...d\vec{p}_N \mbox{~~~~and~each~} d\vec{p}=dp_xdp_ydp_z}\end{displaymath}
\begin{displaymath}{dQ=d\vec{r}_1d\vec{r}_2...d\vec{r}_N \mbox{~~~~and~each~} d\vec{r}=dxdydz\mbox{~~~~}}\end{displaymath}
Writing the Hamiltonian, H in terms of its component kinetic and potential energies gives
\begin{equation}{H = KE + V + V_{ext} = \sum_{i=1}^N\left(\frac{|\vec{p}_i|^2}{2m}\right)+V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)+V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{equation}
where the translational kinetic energy of the N atoms is given by 
\begin{equation}{\text{KE}=\sum_{i=1}^N\left(\frac{|\vec{p}_i|^2}{2m}\right)}\end{equation}
and the potential energy of the total interaction of each atom with the other 
N-1 atoms is given by
\begin{displaymath}{V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{displaymath} 
and where \begin{displaymath}{V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{displaymath} 
is an externally applied potential that varies spatially, such that its 
effects on each atom depend on the atom's position. Often $V_{ext}$ is 
set to zero, as it is in this paper.

Noting that the kinetic energy depends only on momentum while the total 
interatomic and external potential energies depend only on position, 
the integral in Equation~\ref{Z_total} can be separated into two parts 
\begin{equation}{Z=\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}~e^\frac{-\left(\frac{|\vec{p}_1|^2}{2m}+\frac{|\vec{p}_2|^2}{2m}+...\frac{|\vec{p}_N|^2}{2m}\right)}{k_BT}\int{dQ}~e^\frac{-V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)-V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}{k_BT}}\end{equation}  
Multiplying by $\frac{V^N}{V^N}$ where V is the real space volume, and 
\begin{equation}V^N=\int dQ\end{equation} 
the partitian function can then be expressed as
\begin{equation}\label{Zmultiply}{Z=Z_{ideal}Z_{excess}}\end{equation}
where
\begin{align} \label{eq:Zideal}
    Z_{ideal}=V^N\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}~e^\frac{-\left(\frac{|\vec{p}_1|^2}{2m}+ \frac{|\vec{p}_2|^2}{2m}+...\frac{|\vec{p}_N|^2}{2m}\right)}{k_BT}
\end{align}
\begin{align} \label{eq:Zexcess}
    Z_{excess}=\frac{1}{V^N}\int{dQ}~e^\frac{-V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)-V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N )}{k_BT}
\end{align} 
The ideal partitian function $Z_{ideal}$ is that of an ideal gas 
which exhibits no interaction between the atoms - a condition approached 
by a gas at low density where the atoms rarely encounter one another. 
The excess partitian function $Z_{excess}$ accounts for the interaction 
of all the atoms according to their interatomic potential.

@@

                &\approx k_BT\int n_0n_3 + n_1n_2 - \vec n_1\cdot\vec n_2~d\vec u  \label{Fexlowden-inhomo}  \\                
                &\approx k_BT\int \Phi_1 + \Phi_2 + \Phi_3~d\vec u  \\
                &\approx k_BT\int \Phi(\{n_\alpha(\vec u)\})~d\vec u  \\
                &\text{Or, replacing u notation with r} \nonumber \\
                &\approx k_BT\int \Phi(\{n_\alpha(\vec r)\})~d\vec r  \\
                &\approx k_BT\int \Phi(\vec r)~d\vec r             

@@

The FMT equations shown so far have been for a hard-sphere fluid~\cite{Hansen}.
All of the weight functions can be expressed in terms of 
$w_{2}(r)$ which, for a hard-sphere fluid, consists of a delta function.
The derivative of the Mayer function provides 
another important relation expressed in terms of $w_{2}(r)$ 
%that will be important 
%to keep in tact when moving to SFMT~\cite{schmidt2000fluid}.
\begin{align}\label{mayerder=w2convolution}
     \frac{df(r)}{dr} = w_2 \otimes w_2 = \int{w_2(r')w_2(r-r')dr'}
\end{align} 
The weight function $w_{2}(r)$ is in turn related to $w_{3}(r)$ through 
the relation
\begin{align}\label{w2_w3_relation}
    w_{3}=\int_{r}^{\infty}{w_{2}(\vec{r})dr}\mbox{~~~~or equivalently,~~~~}-\frac{\partial{w_3(r)}}{\partial{r}}=w_2(r)
\end{align}
%which will be important to keep in tact when moving to SFMT~\cite{schmidt1999density}.
Thus, all the weight functions, and the above relation are determined
by $w_{3}(r)$.
Soft Fundamental Measure Theory extends FMT to fluids 
with soft potentials by keeping the relationships between the weight 
functions, and Equation~\ref{mayerder=w2convolution} 
%for the derivative of the Mayer function 
the same, but replacing $w_{3}(r)$ with a weight function appropriate 
for a soft-sphere potential~\cite{schmidt1999density, schmidt2000fluid}. 
This is what we do in forming a functional for the Weeks-Chandler-Anderson 
fluid as detailed in the next chapter on Methods.

@@

The parameter $\alpha(T)$ is derived by setting 
the error function potential $V_{erf}(r)$ 
equal to the WCA potential given by Equation~\ref{eq:VWCA} at $r=\alpha$. 
\begin{align}
    V_{erf}(\alpha)=V_{WCA}(\alpha)
\end{align}
This gives
\begin{equation}\label{alphaT}\alpha(T)=\sigma\left(\frac{2}{1+\sqrt{\frac{k_BT}{\epsilon}\ln(2)}}\right)^\frac{1}{6}\end{equation} 
Figure \ref{fig:alphaXivsT} shows how $\alpha(T)$ varies with temperature 
when $\sigma=1$ and $\epsilon=1$.

>>>>

A description of cDFT often begins with considering the Grand Free 
Energy defined as
\begin{equation}\Omega=F-\mu{N}\end{equation}
which can then be expressed in terms of a varying number density 
profile $n(\vec r)$
\begin{equation}\label{GrandFE}\Omega= F_{int} +\int n(\vec{r})\phi_{ext}{(\vec r)}d\vec{r}-\mu\int n(\vec r)d\vec{r}\end{equation}
where
\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation}%
and the intrinsic Helmholtz free energy is defined as
\begin{equation}F_{int} = F - \int n(\vec{r})\phi_{ext}{(\vec r)}d\vec{r}\end{equation}  
%The quantity $\phi_{ext}$
where $\phi_{ext}=\frac{V_{ext}}{N}$ and $V_{ext}(\vec r)$
is the externally applied potential energy given in Equation~\ref{Hamiltonian}. 
The Grand free energy is applicable to systems with a varying number of 
particles and fixed chemical potential, temperature and volume. But in 
this paper a system with a fixed temperature, volume, and number of 
particles will be of interest in which case
%\begin{equation}\int n(\vec r)d\vec{r}=N\end{equation} and so
the Grand free energy is constrained
\begin{equation}\label{GrandFE}\Omega_{constrained}= F[n(\vec r)]-\mu N\end{equation}
The dependence on $n(\vec r)$ is now only present in the expression for 
the Helmholtz free energy, and so it is the Helmholtz energy as a functional 
of the number density profile $F[n(\vec r)]$ that is minimized at equilibrium. 
More specifically, in this paper, the Helmholtz free energy per particle, 
or atom,  $f_N=\frac{F}{N}$ will be minimized which is given by
\begin{equation}f_N=\frac{U}{N}-\frac{TS}{N}\end{equation}
with differential
\begin{equation}\label{usetoshowmin}df_N=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
%This result shows that the natural variables of the Helmholtz free energy 
%per particle $f_N$ are temperature, volume and one over the number of 
%particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. 
%\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}
%Taking the differential
which can be set equal to 
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}\bigg|_{V,\frac{1}{N}}dT 
+\frac{\partial f_N}{V}\bigg|_{T,\frac{1}{N}}dV + \frac{\partial f_N}
{\partial \left(\frac{1}{N}\right)}\bigg|_{T,V}d\left(\frac{1}{N}\right)\end{displaymath}
Using the relation
\begin{displaymath}d\left(\frac{1}{n}\right)=Vd\left(\frac{1}{N}\right)+\frac{1}{N}dV\end{displaymath}
we get
\begin{displaymath}df_N=\frac{\partial f_N}{\partial T}\bigg|_{\frac{1}{n}}dT 
+ \frac{\partial f_N}{\partial \frac{1}{n}}\bigg|_{T}d\left(\frac{1}{n}\right)\end{displaymath}
\begin{equation}f_N(T,\frac{1}{n})\end{equation}
This result shows that temperature and the inverse number density, 
or volume per particle, are natural variables of the Helmholtz free energy 
per particle $f_N$.
%which will thus be minimized when these natural variables are held fixed.

@@

The Helmholtz free energy can be written as
\begin{equation}{F=-k_{B}T\ln(Z)}\end{equation}
which results from equating the expressions for the Helmholtz free energy 
from thermodynamics to those of statistical mechanics. The paritian function 
Z is integrated over all of phase space for N identical particles each with 
a momentum $\vec{p}$ and a position $\vec{r}$, and is given by
\begin{align}  \label{Z_total}
     Z=\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}\int{dQ}~e^\frac{-H(\vec{p}_1,\vec{p}_2,...\vec{p}_N,\vec{r}_1, \vec{r}_2,...\vec{r}_N)}{k_BT}
\end{align}
where
\begin{displaymath}{dP=d\vec{p}_1d\vec{p}_2...d\vec{p}_N \mbox{~~~~and~each~} d\vec{p}=dp_xdp_ydp_z}\end{displaymath}
\begin{displaymath}{dQ=d\vec{r}_1d\vec{r}_2...d\vec{r}_N \mbox{~~~~and~each~} d\vec{r}=dxdydz\mbox{~~~~}}\end{displaymath}
Writing the Hamiltonian, H in terms of its component kinetic and potential energies gives
\begin{equation}\label{Hamiltonian}{H = KE + V + V_{ext} = \sum_{i=1}^N\left(\frac{|\vec{p}_i|^2}{2m}\right)+V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)+V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{equation}
where the translational kinetic energy of the N atoms is given by 
\begin{equation}{\text{KE}=\sum_{i=1}^N\left(\frac{|\vec{p}_i|^2}{2m}\right)}\end{equation}
and the potential energy of the total interaction of each atom with the other 
N-1 atoms is given by
\begin{displaymath}{V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{displaymath} 
and where \begin{displaymath}{V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}\end{displaymath} 
is an externally applied potential that varies spatially, such that its 
effects on each atom depend on the atom's position. Often $V_{ext}$ is 
set to zero, as it is in this paper.

Noting that the kinetic energy depends only on momentum while the total 
interatomic and external potential energies depend only on position, 
the integral in Equation~\ref{Z_total} can be separated into two parts 
\begin{equation}{Z=\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}~e^\frac{-\left(\frac{|\vec{p}_1|^2}{2m}+\frac{|\vec{p}_2|^2}{2m}+...\frac{|\vec{p}_N|^2}{2m}\right)}{k_BT}\int{dQ}~e^\frac{-V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)-V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)}{k_BT}}\end{equation}  
Multiplying by $\frac{V^N}{V^N}$ where V is the real space volume, and 
\begin{equation}V^N=\int dQ\end{equation} 
the partitian function can then be expressed as
\begin{equation}\label{Zmultiply}{Z=Z_{ideal}Z_{excess}}\end{equation}
where
\begin{align} \label{eq:Zideal}
    Z_{ideal}=V^N\frac{1}{N!}\frac{1}{h^{3N}}\int{dP}~e^\frac{-\left(\frac{|\vec{p}_1|^2}{2m}+ \frac{|\vec{p}_2|^2}{2m}+...\frac{|\vec{p}_N|^2}{2m}\right)}{k_BT}
\end{align}
\begin{align} \label{eq:Zexcess}
    Z_{excess}=\frac{1}{V^N}\int{dQ}~e^\frac{-V(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N)-V_{ext}(\vec{r}_1,\vec{r}_2,{...},\vec{r}_N )}{k_BT}
\end{align} 
The ideal partitian function $Z_{ideal}$ is that of an ideal gas.
%an ideal gas which exhibits no interaction between the atoms - a condition approached 
%by a gas at low density where the atoms rarely encounter one another. 
The excess partitian function $Z_{excess}$ accounts for the interaction 
of all the atoms according to their interatomic potential.

@@

                &\approx k_BT\int n_0n_3 + n_1n_2 - \vec n_1\cdot\vec n_2~d\vec u  \label{Fexlowden-inhomo}                  
                %&\approx k_BT\int \Phi_1 + \Phi_2 + \Phi_3~d\vec u  \\
                %&\approx k_BT\int \Phi(\{n_\alpha(\vec u)\})~d\vec u  \\
                %&\text{Or, replacing u notation with r} \nonumber \\
                %&\approx k_BT\int \Phi(\{n_\alpha(\vec r)\})~d\vec r  \\
                %&\approx k_BT\int \Phi(\vec r)~d\vec r             

@@

The FMT equations shown so far have been for a hard-sphere fluid~\cite{Hansen}.
To model a fluid with "soft" spheres which have an interatomic potential 
that decreases smoothly from infinity over a distance of 2R rather than instanteously 
at 2R like that of hard spheres, Schmidt noted that 
all of the weight functions can be expressed in terms of 
$w_{2}(r)$ which, for a hard-sphere fluid, consists of a delta function.
He also noted that the derivative of the Mayer function 
%provides another important relation 
can be expressed in terms of $w_{2}(r)$, 
\begin{align}\label{mayerder=w2convolution}
     \frac{df(r)}{dr} = w_2 \otimes w_2(r) = \int{w_2(r')w_2(r-r')dr'}
\end{align} 
and that the weight function $w_{2}(r)$ is in turn related to $w_{3}(r)$ through 
the relation
\begin{align}\label{w2_w3_relation}
    w_{3}=\int_{r}^{\infty}{w_{2}(\vec{r})dr}\mbox{~~~~or equivalently,~~~~}-\frac{\partial{w_3(r)}}{\partial{r}}=w_2(r)
\end{align}
%which will be important to keep in tact when moving to SFMT~\cite{schmidt1999density}.
%Thus, all the weight functions, and the above relation are determined by $w_{3}(r)$.
Schmidt developed Soft Fundamental Measure Theory (SFMT) which extends FMT to fluids 
with soft potentials by keeping the relationships between the weight 
functions, and Equation~\ref{mayerder=w2convolution} 
%for the derivative of the Mayer function 
the same, but replacing $w_{3}(r)$ with a weight function appropriate 
for a soft-sphere potential~\cite{schmidt1999density, schmidt2000fluid}. 
This is what we do in forming a functional for the Weeks-Chandler-Anderson 
fluid as detailed in the next chapter on Methods.

@@

Unlike the WCA potential, the error function potential $V_{erf}(r)$ is temperature dependent.
To compensate for this temperature dependence so that our functional better approximates
a WCA potential, we make the parameters $\alpha$ and $\Xi$ functions of temperature as well. 
The parameter $\alpha(T)$ is derived by setting 
the error function potential $V_{erf}(r)$ 
equal to the WCA potential given by Equation~\ref{eq:VWCA} at $r=\alpha$, since 
the derivative of the mayer function peaks at this value. 
\begin{align}
    V_{erf}(\alpha)=V_{WCA}(\alpha)
\end{align}
This gives
\begin{equation}\label{alphaT}\alpha(T)=\sigma\left(\frac{2}{1+\sqrt{\frac{k_BT}{\epsilon}\ln(2)}}\right)^\frac{1}{6}\end{equation} 
Figure \ref{fig:alphaXivsT} shows how $\alpha(T)$ varies with temperature 
when $\sigma=1$ and $\epsilon=1$.
